Comparing Models of Delta-Notch Signalling

### 2025-04-11 paper/paper.tex
Revise paper, add Dora's discussion section.

@@

The Delta-Notch signalling pathway plays a crucial role in determining the fate of initially homogeneous cells in a wide array of biological systems.
In this paper, we present a simple yet biologically realistic representation of Delta-Notch signalling and test it using three models: ODE, SDE, and agent-based.
Using the Kolmogorov forward equations, we show that the agent-based model converges to the ODE model in expectation.
We also demonstrate that the SDE and agent-based models can produce cell differentiation even from identical initial conditions.
Finally, we study the effect of parameters such as stochastic noise, domain size, and boundary conditions on the behaviour of our models.
Our findings provide insight into how modelling formalisms can affect simulations of the Delta-Notch signalling network with significant implications for future research.

@@

Cell differentiation is essential for the formation of structure in complex organisms.
But surprisingly, only a small number of signalling pathways are responsible for this important function \cite{bray_notch_2006}.
One of these pathways is the Delta-Notch signalling network, which has been shown to affect neurogenesis in vertebrates by determining which cells adopt neural or glial fates. 
Furthermore, mutations in Notch activity has been shown to be a contributing factor to many cancers \cite{bray_notch_2006}.
The Delta-Notch signalling network is unique because its ligands Delta and Jagged are transmembrane proteins.
This means that Delta-Notch signalling is largely restricted to neighbouring cells. 
However, new findings suggest that cellular protrusions can allow a small fraction of Delta-Notch signals to travel longer distances \cite{berkemeier_coupling_2023}.

\medskip

% TODO: Insert a figure here. The paragraph below should be a caption.

\medskip

Binding of the Delta ligand from cell $B$ to the Notch receptor on cell $A$ results in the \emph{cleavage} (removal) of both the ligand and receptor \cite{bray_notch_2006}.
This interaction triggers the release of NICD (Notch Intracellular Domain) in cell $A$, which \emph{promotes} Notch and \emph{inhibits} Delta.
It is also possible for a Delta ligand to bind to a Notch receptor on the same cell. This phenomenon, known as \emph{cis-inhibition}, cleaves the Notch receptor but does not release NICD.
Serrate (Jagged) is a ligand similar to Delta which can also bind to Notch.
However, NICD promotes Serrate, which creates a double-positive feedaback loop.

\medskip

Previous models of Delta-Notch signalling \cite{collier_pattern_1996, boareto_jaggeddelta_2015} have made a number of assumptions which we also use for our models.
First, cells interact through Delta-Notch signalling if and only if they are adjacent.
Production of Notch is an increasing hill function of NICD.
Production of Delta is a decreasing hill function of NICD.
Both Notch and Delta decay at a rate proportional to their concentration.
Low levels of Notch cause the primary cell fate, while high levels cause the secondary cell fate.
The behaviour of cells with a specific fate depends on the organism and organ, which is outside the scope of these models.

\medskip

The model of Delta-Notch signalling proposed by Collier et al. \cite{collier_pattern_1996} makes other simplifications which violate the modern understanding of the signalling pathway.
These simplifications make it difficult to identify how individual chemical reactions are captured by the model.
For instance, production of activated Notch (NICD) is exclusively a function of the Delta concentration in neighbouring cells, with no regard for the concentration of Notch receptors.
This violates the mass-balance equations, since the quantity of Notch receptors changes dynamically in response to the concentration of Delta ligands in nearby cells.
There are smaller issues, like the absence of cis-inhibition and the Serrate (Jagged)
protein, but these omissions are not necessary to capture the behaviour of the
system.

\medskip

In this paper, we develop three models of the Delta-Notch signalling network based on a modern understanding of the biochemistry of the system. 
Our first model is an agent-based simulation of individual Delta, Notch, and NICD molecules. 
While this model precisely simulates the reaction kinetics under our assumptions, it is impossible to analyze analytically and slow to simulate computationally.
Then, we define two models based on differential equations. 
The ordinary differential equation (ODE) model is equivalent the expected behaviour of the agent-based model (see Appendix \ref{sec:theoretical-justification} for a proof).
The stochastic differential equation (SDE) model is an extension to the ODE model which contains additional noise terms.
These terms are intended to simulate the stochasticity of the agent-based model.
We compare these models with each other using a combination of analytical and computational methods in a variety of different domains.
Furthermore, we show that our models qualitatively replicate the results of Collier et al. \cite{collier_pattern_1996}. 
In particular we show that if a linear array of cells is given zero-flux boundary conditions, the pattern of alternating cell fates arises on the outside edges before gradually spreading inwards towards the center of the domain.
Additionally, both our model and the Collier et al. model exhibit different ratios of primary to secondary fated cells depending on the parameters on a two-dimensional hexagonal lattice.

@@

From the literature \cite{collier_pattern_1996, boareto_jaggeddelta_2015}, we assume Notch ($N$) receptor production has the form $H^{+}(I) = n_{m}I^2/(n_{0}^2 + I^2)$ where $I$ is the NICD concentration and $n_{m}, n_{0}$ are hill function parameters.
Similarly, Delta ligand ($D$) production is governed by a decreasing hill function of the for production has the form $H^{-}(I) = d_{m}d_{0}^2/(d_{0}^2 + I^2)$.
Furthermore, we will assume that Notch and Delta both have decay rate $\gamma$.
The variable $k_{T}$ will denote the binding rate between Notch receptors and Delta ligands.
$\gamma_{I}$ is the decay rate of NICD.
We will also assume that there is no cis-inhibition and no Serrate ligand.

\begin{table}[!htp]
\centering
\begin{tabular}{|m{8em}|m{5em}|m{20em}|} 
 \hline
 Reaction & Rate & Description \\ 
 \hline
 $N_{A} + D_{B} \rightarrow I_{A}$ & 
 $k_{T} N_{A}D_{B}$ &
 A Notch receptor from cell $A$ binds to a delta ligand from cell $B$, triggering the release of NICD in cell $A$. \\
 \hline
 $N_{A} \rightarrow \emptyset$ & 
 $\gamma N_{A}$ & 
 A Notch receptor from cell $A$ decays. \\
 \hline
 $D_{A} \rightarrow \emptyset$ & 
 $\gamma D_{A}$ & 
 A Delta receptor from cell $A$ decays. \\
 \hline
 $I_{A} \rightarrow \emptyset$ &
 $\gamma_{I} I_{A}$ &
 A NICD molecule from cell $A$ decays.  \\
 \hline
 $\emptyset \rightarrow N_{A}$ & 
 $H^{+}(I_{A})$ &
 A Notch receptor in cell $A$ is produced \\
 \hline
 $\emptyset \rightarrow D_{A}$ &
 $H^{-}(I_{A})$ &
 A Delta ligand in cell $A$ is produced \\
 \hline
\end{tabular}
\caption{
  A list of possible reactions and their rates in a cell $A$ with neighbour $B$.
  Since there are $6$ possible reactions per cell, a system with $n$ cells will have $6n$ possible reactions.
  \textbf{Note:} In the Reaction column, $N_A$, $D_B$, and $I_A$ represent one Notch, Delta, and NICD, respectively. In the Rate column, $N_A$, $D_B$, and $I_A$ refer to the number of Notch molecules in cell A, Delta in cell B, and NICD in cell A, respectively.
}
\label{tb:reactions}
\end{table}

\medskip

We will assume that reaction times follow a Poisson point process.
Suppose $E$ is the set of all possible reaction events (the $6$ reactions above for all cells).
Furthermore, let $|E| = N$ be the number of possible reactions.
For each event, let $X_{i} \sim \text{Exp}(\lambda_{i})$ denote the time until the next event of type $i$, where $1 \leq i \leq N$.
Then, the waiting time $T$ in between any two events is:

$$
T = \text{min} \{ X_{1}, \dots, X_{N} \} \sim \text{Exp}\left( \sum_{i = 1}^{N} \lambda_{i} \right) 
$$

This results follows from the definition of an exponential distribution and the memoryless property.
We can use Gillespie's algorithm to simulate the system.
This is the \emph{agent-based} model.
Details of the computational implementation can be found in Appendix \ref{sec:gillespie}.

\medskip

In expectation, the agent-based model converges to the system of ordinary differential equations shown in Equation \ref{eq:deterministic}.
A proof of this fact can be found in the appendix.
This is the \emph{ODE} model, which is described below for a single cell in a two-cell system.
However, this model can be extended to arbitrary domains by defining equations $dN_{i}/dt$, $dD_{i}/dt$, and $dI_{i}/dt$ for each $i \in \{ 1, \dots, n \}$, where $n$ is the number of cells.
In a domain of arbitrary size, $N_{ext}$ and $D_{ext}$ denote the average Notch and Delta concentration over all neighbours.

@@

This model is a simplification of a model by Boareto et al. \cite{boareto_jaggeddelta_2015}.
However, to more directly compare our results with those of Collier et al. \cite{collier_pattern_1996}, we chose to ignore Serrate and cis-inhibition.
Adding NICD, which was absent from Collier et al., was necessary for the assignment of probabilities to each reaction event.
An analysis of the system of ODEs in Equation \ref{eq:deterministic} for fixed values of $N_{ext}, D_{ext}$ reveals an $N = 0$ steady state (the \emph{primary fate}) for all parameter values.
There is also a high Notch steady state (the \emph{secondary fate}) which emerges from a fold bifurcation when $D_{ext}$ is sufficiently large.
See Appendix \ref{sec:bifurcation} for more information.

\medskip

To capture the inherent randomness in receptor-ligand interactions while maintaining computational efficiency, we implement a stochastic differential equation \emph{SDE} model based on the Langevin equation formalism.
This approach represents an intermediate level of detail between the ODE model and the fully discrete agent-based model.
Starting from the ODE model described in Equation \ref{eq:deterministic}, we incorporate stochastic fluctuations by adding noise terms proportional to the square root of the reaction rates to get Equation \ref{eq:stochastic}.

@@

In Equation \ref{eq:stochastic}, $dW_N$, $dW_D$, and $dW_I$ are independent Wiener processes representing white noise.
The structure of the noise terms follows from the assumption that noise in each reaction channel is proportional to the current concentration of those molecules.

\medskip

To implement these models computationally, we employed the Euler-Maruyama method with time step $dt$.
The deterministic increment calculated first, and then the Wiener increments are sampled from a normal distribution with mean $0$ and standard deviation $\sqrt{dt}$.
Then, each Wiener increment is scaled by its respective variable and added to the deterministic increment.

@@

% TODO: Write some filler text to connect the different experiments.

@@

\medskip

@@

\medskip

When we defined the SDE model in Equation \ref{eq:stochastic}, we introduced a parameter $\sigma$ that represents the amount of noise.
However, with the methods we know, it is difficult to estimate the value of this parameter analytically or numerically.
Here, we present a computational method to get a qualitative estimate of $\sigma$ which uses the differentiation time of the two-cell system.
We define the differentiation time $t^{*}$ to be the time at which one cell adopts the primary fate ($N < 1$).
Through our experiments, we noticed that higher levels of noise lower the average differentiation time by rapidly perturbing the system into more ``extreme'' states.
Using this relationship, we estimated that $\sigma = 0.02$ by comparing simulations of the SDE model with the agent-based model.
See Figure \ref{fig:noise-estimate} for a summary of this result.

@@

\begin{figure}[!htp]
   \makebox[\textwidth][c]{\includegraphics[width=1.3\textwidth]{img/vis13.png}}
   \caption{Results of two-at-a-time stability analysis of the deterministic, stochastic, and agent-based modelling formalisms on a two-cell domain with periodic boundary conditions. The regions enclosed by the red, blue, and green lines are regions in which cell differentiation occurs for each of the three models. Regions were obtained by simulating a $25 \times 25$ grid of test points and taking the convex hull of the points at which cell differentiation occurred. This analysis suggests that the deterministic and stochastic ODE models are on average less sensitive to changes in parameter values than the agent-based model. In particular, the agent-based model does not achieve cell differentiation for large values of the Notch and Delta production parameters $n_{m}$ and $d_{m}$.} 
   \label{fig:stability-analysis}
\end{figure}

@@

A linear domain is an array of cells in which each cell receives and emits Delta-Notch signals to $2$ neighbours ($1$ if the cell is on the boundary of domain).
Collier et al. \cite{collier_pattern_1996} tested their model on linear domains of up to $40$ cells with zero-flux boundary conditions.
They observed that cells on the boundary of the domain adopted the primary fate first.
Then, the alternating pattern of primary and secondary fate cells gradually spread inwards towards the center of the domain.
In our first experiment on a linear domain, we replicated this result as shown in Figure.

@@

% TODO: Write this section. Should only need 2-3 paragraphs.

@@

Plotting the nullclines in Equations \ref{eq:one-cell-notch-nullcline} and \ref{eq:one-cell-nicd-nullcline} on the $N-I$ plane for small and large values of $D_{ext}$ (assuming the other parameters are suitably chosen) reveals a fold bifurcation. This is shown in Figure \ref{fig:one-cell-phase-plane}.

>>>>

\usepackage{setspace}
\doublespacing

\usepackage{ragged2e}
\justifying

@@

The Delta-Notch signalling pathway plays a crucial role in driving the differentiation of initially homogeneous cells in a variety of biological systems.
In this paper, we present a simple biological model of Delta-Notch signalling and investigate it using three mathematical formalisms: ODE, SDE, and agent-based.
Using the Kolmogorov forward equations, we show that the agent-based model converges to the ODE model in expectation.
We also demonstrate that the SDE and agent-based models can produce cell differentiation even from identical initial conditions.
Finally, we study the effect of parameters such as stochastic noise, domain shape, and boundary conditions on the behaviour of our models.
Our findings provide insight into how modelling formalisms can affect simulations of the Delta-Notch signalling network with significant implications for future research.

@@

Cell differentiation is essential for the formation of structure in complex organisms.
However, only a few signalling pathways are responsible for this important function \cite{bray_notch_2006}.
One of these pathways is the Delta-Notch signalling network, which has been shown to drive neurogenesis in vertebrates by determining which cells adopt neural and glial fates. 
Furthermore, mutations in Notch activity has been shown to be a contributing factor to many cancers \cite{bray_notch_2006}.
The Delta-Notch signalling network is unique because its ligands Delta and Jagged are transmembrane proteins.
This means that Delta-Notch signalling is largely restricted to neighbouring cells. 
However, new findings suggest that cellular protrusions can allow a small fraction of Delta-Notch signals to travel longer distances \cite{berkemeier_coupling_2023}.

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{img/signalling-network.png}
  \caption{
Illustration of the Delta-Notch signalling network.
Binding of the Delta and Jagged ligands from cell $B$ to the Notch receptors on cell $A$ results in the \emph{cleavage} (removal) of both the ligand and receptor \cite{bray_notch_2006}.
This interaction triggers the release of NICD (Notch Intracellular Domain) in cell $A$, which \emph{promotes} Notch, \emph{inhibits} Delta, and \emph{promotes} Jagged.
It is also possible for a Delta ligand to bind to a Notch receptor on the same cell. This phenomenon, known as \emph{cis-inhibition}, cleaves the Notch receptor but does not release NICD.
  }
\end{figure}

Interactions between Delta, Notch, and Jagged on the cell membrane cause some cells to adopt a low-Notch state (the \emph{primary fate}) and others to adopt a high-Notch state (the \emph{secondary fate}).
Typically, primary and secondary fated cells form an alternating pattern.
Then, through other signalling networks, the primary and secondary fated cells adopt different characteristics depending on the organism and organ.

The model of Delta-Notch signalling proposed by Collier et al. \cite{collier_pattern_1996} is one of the seminal works in the field.
Through their ODE model, Collier et al. show that if a linear array of cells is given zero-flux boundary conditions, a pattern of alternating cell fates arises on the outside edges before gradually spreading inwards towards the center of the domain.
Additionally, Collier et al. show that their model exhibits different ratios of primary to secondary fated cells depending on the parameters when simulated on a two-dimensional hexagonal lattice. 
While these qualitative results are important, Collier et al. use many simplifications that make it difficult to identify how individual chemical reactions are captured by the model.
For instance, production of activated Notch (NICD) is exclusively a function of the Delta concentration in neighbouring cells, without regard for the concentration of Notch receptors.
This violates the mass-balance equations, since we now know that the quantity of Notch receptors changes dynamically in response to the concentration of Delta ligands in nearby cells.
The ODE model of Delta-Notch signalling proposed by Boareto et al. \cite{boareto_jaggeddelta_2015} is built on a more modern understanding of the biochemistry of the system.
Their model respects the mass-balance equations and explicitly includes binding rate and decay rates. 
The Boareto et al. model also assumes that production of Notch is an increasing hill function of NICD while production of Delta is a decreasing hill function of NICD, a useful simplification that we will also use.
However, Boareto et al. \cite{boareto_jaggeddelta_2015} only analyze their model on a two-cell domain, which makes it difficult to generalize their results to larger groups of cells.

In this paper, we develop three models of the Delta-Notch signalling network. 
Our first model is an agent-based simulation of individual Delta, Notch, and NICD molecules. 
While this model precisely simulates the reaction kinetics, it is impossible to analyze analytically and slow to simulate computationally.
Then, we define two models based on differential equations.
The ODE model is a deterministic representation of the approximate reaction kinetics, while the SDE model includes noise terms to better replicate the agent-based model.
To connect our agent-based and ODE models, we use the Kolmogorov forward equations, taking their expectation and applying the law of total expectation to derive a system of ODEs for the expected molecular counts (see Appendix \ref{sec:kfe}, \ref{sec:theoretical-justification}).
This calculation relies on the assumption that the propensity functions are approximately linear or that fluctuations around the mean are small, which holds when molecule numbers are large.
Under this assumption, evaluating the propensity functions at the expected state recovers our deterministic ODE model after appropriate scaling.
We then compare all three models with each other using a combination of analytical and computational methods in a variety of different domains.
Furthermore, we show that our models qualitatively replicate the results of Collier et al. \cite{collier_pattern_1996}.

@@

From the literature \cite{collier_pattern_1996, boareto_jaggeddelta_2015}, we assume Notch ($N$) receptor production has the form $H^{+}(I) = n_{m}I^2/(n_{0}^2 + I^2)$ where $I$ is the NICD concentration and $n_{m}, n_{0}$ are hill function parameters.
Similarly, we assume production of the Delta ligand ($D$) is governed by a decreasing hill function of the form $H^{-}(I) = d_{m}d_{0}^2/(d_{0}^2 + I^2)$.
Furthermore, we will assume that Notch and Delta both have decay rate $\gamma$.
The variable $k_{T}$ will denote the binding rate between Notch receptors and Delta ligands.
$\gamma_{I}$ is the decay rate of NICD.
We will also assume that there is no cis-inhibition and no Jagged ligand.

\begin{table}[!htp]
\centering
\begin{tabular}{|m{8em}|m{5em}|m{20em}|} 
 \hline
 Reaction & Rate & Description \\ 
 \hline
 $N_{A} + D_{B} \rightarrow I_{A}$ & 
 $k_{T} N_{A}D_{B}$ &
 A Notch receptor from cell $A$ binds to a delta ligand from cell $B$, triggering the release of NICD in cell $A$. \\
 \hline
 $N_{A} \rightarrow \emptyset$ & 
 $\gamma N_{A}$ & 
 A Notch receptor from cell $A$ decays. \\
 \hline
 $D_{A} \rightarrow \emptyset$ & 
 $\gamma D_{A}$ & 
 A Delta receptor from cell $A$ decays. \\
 \hline
 $I_{A} \rightarrow \emptyset$ &
 $\gamma_{I} I_{A}$ &
 A NICD molecule from cell $A$ decays.  \\
 \hline
 $\emptyset \rightarrow N_{A}$ & 
 $H^{+}(I_{A})$ &
 A Notch receptor in cell $A$ is produced \\
 \hline
 $\emptyset \rightarrow D_{A}$ &
 $H^{-}(I_{A})$ &
 A Delta ligand in cell $A$ is produced \\
 \hline
\end{tabular}
\caption{
  A list of possible reactions and their rates in a cell $A$ with neighbour $B$.
  Since there are $6$ possible reactions per cell, a system with $n$ cells will have $6n$ possible reactions.
  In the Reaction column, $N_A$, $D_B$, and $I_A$ represent one Notch, Delta, and NICD, respectively. In the Rate column, $N_A$, $D_B$, and $I_A$ refer to the number of Notch molecules in cell A, Delta in cell B, and NICD in cell A, respectively.
}
\label{tb:reactions}
\end{table}

We will assume that reaction times follow a Poisson point process.
Suppose $E$ is the set of all possible reaction events (the $6$ reactions above for all cells).
Furthermore, let $|E| = N$ be the number of possible reactions.
For each event, let $X_{i} \sim \text{Exp}(\lambda_{i})$ denote the time until the next event of type $i$, where $1 \leq i \leq N$.
Then, the waiting time $T$ in between any two events is given by the formula in Equation \ref{eq:waiting-time} due to the memoryless property.

\begin{equation}
T = \text{min} \{ X_{1}, \dots, X_{N} \} \sim \text{Exp}\left( \sum_{i = 1}^{N} \lambda_{i} \right) 
\label{eq:waiting-time}
\end{equation}

To simulate the individual reactions, we can use Gillespie's algorithm.
This is the agent-based model.
Details of the computational implementation can be found in Appendix \ref{sec:gillespie}.
In expectation, the agent-based model converges to the system of ordinary differential equations shown in Equation \ref{eq:deterministic}.
A proof of this fact can be found in Appendix \ref{sec:theoretical-justification}.
This is the ODE model, which is described below for a single cell in a two-cell system.
However, this model can be extended to arbitrary domains by defining equations $dN_{i}/dt$, $dD_{i}/dt$, and $dI_{i}/dt$ for each $i \in \{ 1, \dots, n \}$, where $n$ is the number of cells.
In a domain of arbitrary size, $N_{ext}$ and $D_{ext}$ are the average Notch and Delta concentrations over all neighbours.

@@

This model is a simplification of the model by Boareto et al. \cite{boareto_jaggeddelta_2015}.
However, to more directly compare our results with those of Collier et al. \cite{collier_pattern_1996}, we chose to ignore Jagged and cis-inhibition.
Adding NICD, which was absent from Collier et al., was necessary for the assignment of probabilities to each reaction event.
An analysis of the system of ODEs in Equation \ref{eq:deterministic} for fixed values of $N_{ext}, D_{ext}$ reveals an $N = 0$ steady state (the \emph{primary fate}) for all parameter values.
There is also a high Notch steady state (the \emph{secondary fate}) which emerges from a fold bifurcation when $D_{ext}$ is sufficiently large.
See Appendix \ref{sec:bifurcation} for more information.

To capture the inherent randomness in receptor-ligand interactions while maintaining computational efficiency, we implement a stochastic differential equation (SDE) model based on the Langevin equation formalism.
This approach represents an intermediate level of detail between the ODE model and the fully discrete agent-based model.
Starting from the ODE model described in Equation \ref{eq:deterministic}, we incorporate stochastic fluctuations by adding noise terms proportional to the square root of the reaction rates to get Equation \ref{eq:stochastic}.

@@

In Equation \ref{eq:stochastic}, $dW_N$, $dW_D$, and $dW_I$ are independent Wiener processes representing white noise.
The structure of the noise terms follows from the assumption that noise in each reaction channel is proportional to the current concentration of those molecules.
To implement the ODE and SDE models computationally, we employed the Euler-Maruyama method with time step $dt$.
The deterministic increment calculated first, and then the Wiener increments are sampled from a normal distribution with mean $0$ and standard deviation $\sqrt{dt}$.
Then, each Wiener increment is scaled by its respective variable and added to the deterministic increment.

@@

@@

@@

When we defined the SDE model in Equation \ref{eq:stochastic}, we introduced a parameter $\sigma$ that represents the amount of noise.
However, with the methods we know, it is difficult to estimate the value of this parameter analytically or numerically.
Here, we present a computational method to get an estimate of $\sigma$ which uses the differentiation time of the two-cell system.
We define the differentiation time $t^{*}$ to be the time at which one cell adopts the primary fate ($N < 1$).
Through our experiments, we noticed that higher levels of noise lower the average differentiation time by rapidly perturbing the system into more ``extreme'' states.
Using this relationship, we estimated that $\sigma = 0.02$ by comparing simulations of the SDE model with the agent-based model.
See Figure \ref{fig:noise-estimate} for a summary of this result.

@@

\begin{figure}[!htp]
   \makebox[\textwidth][c]{\includegraphics[width=1.3\textwidth]{img/vis13.png}}
   \caption{Results of two-at-a-time stability analysis of the deterministic, stochastic, and agent-based modelling formalisms on a two-cell domain with periodic boundary conditions. The regions enclosed by the red, blue, and green lines are regions in which cell differentiation occurs for each of the three models. Regions were obtained by simulating a $25 \times 25$ grid of test points and taking the convex hull of the points at which cell differentiation occurred. The regions are plotted with logarithmic axes. This analysis suggests that the deterministic ODE model is most sensitive to changes in the parameter regime, especially changes to $\gamma$ and $\gamma_{I}$.} 
   \label{fig:stability-analysis}
\end{figure}

@@

Collier et al. \cite{collier_pattern_1996} tested their model on linear domains of up to $40$ cells with zero-flux boundary conditions.
They observed that cells on the boundary of the domain adopted the primary fate first.
Then, the alternating pattern of primary and secondary fated cells gradually spread inwards towards the center of the domain.
In our first experiment on a linear domain, we replicated this result as shown in Figure.

@@

Our comparison of the ODE, SDE, and agent-based models highlights a clear trade-off between simplicity, accuracy to the biology, and the role of stochasticity in modelling Delta-Notch signalling.
The ODE model is computationally efficient and easy to analyze, but it can only break symmetry with an initial perturbation.
In other words, stable cell differentiation only occurs when the initial conditions are already imbalanced.
The SDE model, by introducing intrinsic noise, allows cells to break symmetry without perturbations.
This behaviour closely mirrors the agent-based model, which captures randomness at the level of individual molecules and allows differentiation to emerge purely from fluctuations in the receptor-ligand dynamics.

Simulations of the agent-based model revealed that it is difficult to control cell fate using initial state alone.
Even when we imposed large initial perturbations in the two-cell system ($\pm 25\%$), the intended outcome only occurred about $80\%$ of the time.
When comparing the SDE model to the agent-based model, we found that the differentiation time varied considerably depending on the noise coefficient $\sigma$.
To ensure these two models produced consistent results, we carefully tuned the noise coefficient to an optimal value of $\sigma = 0.02$.
Finally, we concluded our investigation of the two-cell domain with a stability analysis, which found that the ODE model is the most sensitive to changes in parameter values, especially the decay rates $\gamma$ and $\gamma_{I}$.
On linear domains, all three models exhibited the behaviour described by Collier et al., with an alternating pattern of cell differentiation spreading inward from the edges.
The ODE model generated clean, perfectly alternating patterns, consistent with its deterministic structure.
The SDE model introduced occasional irregularities but largely preserved alternation.
However, the agent-based model produced the most diverse outcomes — especially in small or oddly shaped domains — where patterns often deviated from strict alternation.
This variability points to a key biological insight: while stochasticity provides flexibility and diversity in pattern formation, it can also make patterns less predictable, particularly in small systems or near critical parameter boundaries.

@@

Plotting the nullclines in Equations \ref{eq:one-cell-notch-nullcline} and \ref{eq:one-cell-nicd-nullcline} on the $N-I$ plane for small and large values of $D_{ext}$ (assuming the other parameters are suitably chosen) reveals a fold bifurcation. This is shown in the figure below.

### 2025-04-11 snacks-presentation/main.tex
Revise paper, add Dora's discussion section.

\begin{frame}
  \frametitle{Our Chemical Equations}
  Consider a cell $A$ with neighbour $B$. The Notch, Delta, and NICD concentrations of cell $A$ are given by $N_{A}$, $D_{A}$, and $I_{A}$ respectively (and similarly for cell $B$).
  \begin{enumerate} 
    \item $N_{A} + D_{B} \rightarrow I_{A}$: Binding event. 
    \item $N_{A} \rightarrow \emptyset$: Notch decay.
    \item $D_{A} \rightarrow \emptyset$: Delta decay.
    \item $I_{A} \rightarrow \emptyset$: NICD decay.
    \item $\emptyset \rightarrow N_{A}$: Notch production.
    \item $\emptyset \rightarrow D_{A}$: Delta production.
  \end{enumerate}
\end{frame}

@@

  \frametitle{Our Models}

  \onslide<1->{In the \textbf{Agent-Based} model, we simulate the chemical reactions directly using the \emph{Gillespie Algorithm}. Under our assumptions, this is the most accurate model.}

@@

  \onslide<2->{In the \textbf{ODE} model, we approximate the reaction kinetics by simulating the agent-based model \emph{in expectation}. To ensure cell differentiation, we apply an \emph{initial perturbation} to all simulations.}

@@

  \onslide<3->{In the \textbf{SDE} model, we reintroduce randomness by adding an independent \emph{Wiener Process} to each equation in the ODE model. This recreates some the stochasticity of the agent-based model.}

@@

% NOTE: I can extend this section by talking about the Gillespie algorithm.

@@

  \onslide<1->{
    We tested model stability by perturbing $5$ parameters: $N_{m}$, $D_{m}$, $K_{T}$, $\gamma$, and $\gamma_{I}$.
    \begin{itemize}
      \item $N_{m}$, $D_{m}$ are the maximum rates of Notch/Delta production.
      \item $K_{T}$ is the binding rate.
      \item $\gamma$, $\gamma_{I}$ are the Notch/Delta and NICD decay rates.
    \end{itemize}
  }

@@

\begin{frame}
  \frametitle{Stability Analysis}
\end{frame}

@@

>>>>

@@

  \frametitle{Our Chemical Equations}

  \onslide<1->{Consider a cell $A$ with neighbour $B$. The Notch, Delta, and NICD concentrations of cell $A$ are given by $N_{A}$, $D_{A}$, and $I_{A}$ respectively (and similarly for cell $B$).

@@

  \begin{enumerate} 
    \item $N_{A} + D_{B} \rightarrow I_{A}$: Binding event. 
    \item $N_{A} \rightarrow \emptyset$: Notch decay.
    \item $D_{A} \rightarrow \emptyset$: Delta decay.
    \item $I_{A} \rightarrow \emptyset$: NICD decay.
    \item $\emptyset \rightarrow N_{A}$: Notch production.
    \item $\emptyset \rightarrow D_{A}$: Delta production.
\end{enumerate}}

@@

  \onslide<2->{In the \textbf{Agent-Based} model, we simulate these reactions directly using the \emph{Gillespie Algorithm}. Under our assumptions, this is the most accurate model.}

@@

  \frametitle{Differential Equation Models}

  \onslide<1->{In the \textbf{ODE} model, we approximate the reaction kinetics by simulating the agent-based model \emph{in expectation}. Since this model is deterministic, we ensure cell differentiation with an initial perturbation before all simulations.

$$\begin{aligned}
  \frac{dN}{dt} &= \frac{n_{m}I^2}{n_{0}^2 + I^2} - k_{T}ND_{ext} - \gamma N \\
  \frac{dD}{dt} &= \frac{d_{m}d_{0}^2}{d_{0}^2 + I^2} - k_{T}DN_{ext} - \gamma D \\
  \frac{dI}{dt} &= k_{T}ND_{ext} - \gamma_{I}I
\end{aligned}$$

  }

  \onslide<2->{In the \textbf{SDE} model, we recreate some of the stochasticity of the agent-based model by adding an independent \emph{Wiener Process} to each equation above.}

@@

  \onslide<1->{
    We tested model stability by perturbing $N_{m}$, $D_{m}$, $K_{T}$, $\gamma$, and $\gamma_{I}$.
$$\begin{aligned}
  \frac{dN}{dt} &= \frac{n_{m}I^2}{n_{0}^2 + I^2} - k_{T}ND_{ext} - \gamma N \\
  \frac{dD}{dt} &= \frac{d_{m}d_{0}^2}{d_{0}^2 + I^2} - k_{T}DN_{ext} - \gamma D \\
  \frac{dI}{dt} &= k_{T}ND_{ext} - \gamma_{I}I
\end{aligned}$$
  }

@@

\begin{frame}
  \frametitle{Stability Analysis Results}
\end{frame}

@@

\section{End}
